---
title: "Preliminary Data Visualisation of Auckland AQI Data"
author: "Stephen Su"
output:
  html_document:
    code_folding: hide
    df_print: paged
    fig_caption: yes
    number_sections: no
  css: style.css
---

<style type="text/css">

body {
  font-family: serif;
}

</style>

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  message = FALSE, warning = FALSE, fig.retina = 3, fig.align = "center",
  comment = "#>", dev.args = list(type = "cairo"), dpi = 200
)
```

```{r prep}
invisible(purrr::map(c(
  paste0(
    "../prelim/ggplot2-demo/",
    c(
      "packages", "data-clean", "covid19-lvl",
      "function", "plot"
    ), ".R"
  ),
  paste0(
    "../prelim/echarts4r-demo/",
    c("packages", "function", "plot"),
    ".R"
  )
), source))
```


## The Dataset
The data includes the hourly air quality index (AQI) from 2019 to 2020 for Glen Eden, Henderson, Penrose, Queen Street and Takapuna, downloaded from the [Auckland Council Environmental Data Portal](https://environmentauckland.org.nz/) on 17 March 2021. The dataset consists of 87,720 observations, including missing values, and in addition to AQI, the dataset also includes, for each observation, the NZ COVID-19 alert levels.

The table below shows the *first 10* observations of the Auckland AQI dataset.

```{r data}
aqi_data <- "../data/akl-aqi-19-20.csv" %>%
  read_csv(locale = locale(tz = "Pacific/Auckland")) %>%
  tsibble(index = datetime, key = location)

reactable(head(aqi_data, 10), bordered = TRUE, striped = TRUE, highlight = TRUE)
```

## AQI Level Map
```{r expl, fig.height = 8}
aqi_data %>%
  dplyr::mutate(aqi = aqi_cat(aqi)) %>%
  cat_heats(aqi, aqi_pal) +
  ggplot2::facet_grid(fct_reorder(location, -get_lat(location)) ~ .) +
  theme(legend.position = "top") +
  labs(
    col = "AQI in Auckland",
    x = "Date",
    y = "Hour of Day"
  )

kable(rbind(
  unique(aqi_data[["location"]]),
  unique(aqi_data[["location"]]) %>%
    map_dbl(function(x) {
      round(mean(is.na(filter(aqi_data, location == x)[["aqi"]])) * 100, 2)
    }) %>%
    paste0("% missing")
))
```

## Seasonal Variation
```{r season, fig.height = 4}
gg_botsplot(aqi_data, outlier.shape = 1) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  labs(x = "Month", y = "AQI")

gg_seasquantile(aqi_data) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  labs(
    x = "Month", y = "AQI",
    title = "1~100% Quantile"
  )

gg_seasquantile(aqi_data, q = seq(.1, .9, .001)) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  labs(
    x = "Month", y = "AQI",
    title = "10~90% Quantile"
  )
```
**On 22/10/2019, a fire broke out in the NZ International Convention Centre, which is less than 1km away from the Queen St weather observatory, the nearest AQI station. ** **The fire burnt for two days, unleashing a vast amount of PM2.5 and PM10, causing an extraordinary spike in AQI. Which is shown on both the boxplot and quantile plot. The plot of the 2.5~97.5% AQI quantile indicates a peak in average AQI in mid-Winter.**

## Hourly Variation within One Day
```{r hourly, fig.height = 4}
gg_seasquantile(aqi_data, period = "day") +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
  labs(
    x = "Time of Day", y = "AQI",
    title = "1~100% Quantile"
  )

gg_seasquantile(aqi_data, period = "day", q = seq(.1, .9, .001)) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
  labs(
    x = "Time of Day", y = "AQI",
    title = "10~90% Quantile"
  )
```
**It seems that AQI on Queen Street is the most affected by daily traffic peaks.**

## The Effect of COVID-19 on AQI
```{r lockdown, fig.height = 4}
aqi_data %>%
  ggplot(aes(x = akl_level, y = aqi)) +
  geom_boxplot(outlier.shape = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  labs(
    title = "The Effect of COVID-19 Alert Levels on AQI",
    x = "COVID-19 Alert Level",
    y = "AQI"
  )

aqi_data %>%
  ggplot(aes(x = covid19_period, y = aqi)) +
  geom_boxplot(outlier.shape = 1) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1)) +
  ggplot2::facet_grid(~ fct_reorder(location, -get_lat(location))) +
  ggplot2::scale_y_continuous(trans = "sqrt") +
  labs(
    title = "The Effect of Total Lockdown on AQI",
    x = "COVID-19 Period",
    y = "AQI (Square Root Scale)"
  )

filter(aqi_data, location == "queen_street") %>%
  update_tsibble(key = covid19_period) %>%
  gg_seasquantile(q = seq(.1, .9, .001), period = "day") +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) +
  labs(
    x = "Time of Day", y = "AQI",
    title = "10~90% Quantile of AQI on Queen Street"
  )
```
**It seems that only a total lockdown has an effect significant on AQI. During alert Level 4, the "traffic peaks" in Queen Street perished as the median area was flat.**

## The Effect of Location on AQI
```{r location, fig.height = 8}
aqi_data %>%
  gg_seasquantile(q = seq(.1, .9, .005), polar = TRUE, size = .1) +
  facet_geo(~location, grid = akl_city_grid) +
  labs(
    title = "10~90% Quantile of AQI by Relative Geolocation",
    y = "AQI"
  )
```
